Zero-shot task adaptation by homoiconic meta-mapping

### 2019-09-13 writing/homoiconic_meta_mapping_main.tex
Tweaks

%\vspace{-0.25em}
\section{Homoiconic meta-mapping (HoMM) architecture}
%\vspace{-0.5em} % section
To address these challenges, we propose HoMM architectures, composed of two components: 
%\vspace{-1em}
\begin{enumerate} \setlength \itemsep{0em}
\item Input/output systems: domain specific encoders and decoders (vision, language, etc.) that map into a shared embedding space $Z$.
\item A meta-learning system that a) learns to embed tasks into the shared embedding space $Z$, b) learns to use these task embeddings to perform task-appropriate behavior, c) learns to embed meta-mappings into the same space, and d) learns to use these meta-mapping embeddings to transform basic task embeddings in a meta-mapping appropriate way.\end{enumerate}
%\vspace{-0.75em}
\looseness=-1
These architectures are homoiconic because they have a completely shared $Z$ for individual data points, tasks, and meta-mappings. Why is this useful? The primary advantage is that it parsimoniously allows for arbitrary mappings between these entities. In addition to basic tasks, the system can learn to perform meta-mappings to follow instructions or change behavior. That is, it can transform task representations using the same components it uses to transform basic data points. (See also appendix \ref{app_lesion_results_shared_z}.) \par
Without training on meta-mappings, of course, the system will not be able to execute them well. However, as we will show, if it is trained on a broad enough set of such mappings, it will be able to generalize to new instances drawn from the same meta-mapping distribution. For instances that fall outside its data distribution, or for optimal performance, it may require some retraining, however. This reflects the structure of human behavior -- we are able to adapt rapidly when new knowledge is relatively consistent with our prior knowledge, but learning an entirely new paradigm (such as calculus for a new student) can be quite slow \citep[cf.][]{Kumaran2016, Botvinick2019}. \par 
\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.8, every node/.style={scale=0.8}]
%% basic meta learning
\begin{scope}[shift={(0.4, 0.4)}]
\draw[boundingbox, fill=white] (-3, -4.3) rectangle (2.5, 3.2);
\end{scope}
\begin{scope}[shift={(0.2, 0.2)}]
\draw[boundingbox, fill=white] (-3, -4.3) rectangle (2.5, 3.2);
\end{scope}

@@

\caption{The HoMM architecture allows for transformations at different levels of abstraction. For basic meta-learning (gray), a dataset consisting of (input embedding, output embedding) tuples is processed by the meta-network \(\mathcal{M}\) to produce a function embedding \(z^{func}\), which is processed by the hyper network \(\mathcal{H}\) to parameterize a function \(F_{z^{func}}\), which attempts to compute the transformation on held-out probe inputs. However, our approach goes beyond basic meta-learning. The function embedding \(z^{func}\) can then be seen as a single input or output at the next level of abstraction, when the same networks \(\mathcal{M}\) and \(\mathcal{H}\) are used to transform function embeddings based on examples of a meta-mapping (purple box). To evaluate meta-mapping performance, a probe embedding of a held-out function is transformed by the architecture to yield a predicted embedding for the transformed task. The performance of this predicted embedding is evaluated by moving back down a level of abstraction and evaluating on the actual target task (blue box). Because the function embedding is predicted by a transformation rather than computed from examples, this allows for zero-shot performance of a new task. (\(M\) and \(H\) are learnable deep networks, and $F_{z}$ is a deep network parameterized by $\mathcal{H}$ conditioned on function embedding \(z\). Input and output encoders/decoders are omitted for simplicity. See the text and appendix \ref{app_model_details} for details.)} \label{architecture_inference_fig}

@@

\paragraph{Basic meta-learning:} To perform a basic task, input and target encoders ($\mathcal{I}$ and $\mathcal{T}$) are used to embed individual pairs from an example dataset \(D_1\), to form a dataset of example (input, output) tuples (fig. \ref{architecture_inference_fig}a). These examples are fed to $\mathcal{M}$, which produces a function embedding (via a deep neural network, including an element-wise max across the examples). This function embedding is mapped through the hyper network $\mathcal{H}$ to parameterize $F$, and then $F$ is used to process a dataset of embedded probe inputs, and $\mathcal{O}$ to map the resultant embeddings to outputs. This system can be trained end-to-end on target outputs for the probes. Having two distinct datasets forces generalization at the meta-learning level, see appendix \ref{app_clarifying_holdouts}. See appendix \ref{app_model_details} for detailed architecture, and hyper-parameters. \par 

@@

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/poly/meta_results.png}
\caption{The polynomials domain, section \ref{sec_poly}.}
\label{poly_meta_map_results_examples}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/meta_mapping.png}
\caption{The cards domain, section \ref{sec_cards}.}
\label{cards_meta_map_results_examples}
\end{subfigure}%
\caption{The HoMM architecture performs well at meta-mappings. (\subref{poly_meta_map_results_examples}) The system generalizes to apply learned meta-mappings to new polynomials, and to apply unlearned meta-mappings. The plots show the loss produced when evaluating the mapped embedding on the target task. For example, if the initial polynomial is $p(x) = x + 1$, and the meta-task is ``square,'' the loss would be evaluated by applying this meta-mapping to the embedding of $p(x)$ and evaluating how well the function specified by the mapped embedding regresses $p(x)^2 = x^2 + 2x + 1$. The system succeeds at applying meta-mappings it is trained on to held-out polynomials, as well as applying held-out meta-mappings to either trained or held-out polynomials. The solid line indicates optimal performance; the dashed line is untrained model performance.
(\subref{cards_meta_map_results_examples}) The system generalizes to meta-mapping new tasks in the cards domain. The system is trained to do the meta-mappings shown here on a subset of its basic tasks, and is able to generalize these mappings to perform novel tasks zero-shot. For example, for the ``losers'' mapping, the sytem is trained to map games to their losers variants. When given a held-out game, it is able to apply the mapping to guess how to play the losing variation. This plot shows the reward produced by taking the mapped embedding and playing the targeted game. The gray dashed line indicates random performance, while the colored dashed lines indicate performance if the system did not alter its behavior in response to the meta-mapping. The system generally exceeds these baselines, although the switch-suits baseline is more difficult with the targeted holdout. 
Error-bars are bootstrap 95\%-CIs.} 
\label{meta_map_results}
\end{figure}

As a proof of concept, we first evaluated the system on the task of learning polynomials of degree $\leq 2$ in 4 variables (i.e. the task was to regress functions of the form $p: \mathbb{R}^4 \rightarrow \mathbb{R}$ where $p \in \mathcal{P}_2 \left(\mathbb{R}\right)$, though the model was given no prior inductive bias toward polynomial forms). For example, if $p(w,x,y,z) = x$, the model might see examples like $(-1,1,1,1; 1)$ and $(0.7, 2.1, 1.3, -4; 2.1)$, and be evaluated on its output for points like $(-1, -1.3, 0.5, 0.3)$. This yields an infinite family of base-level tasks (the vector space of all such polynomials), as well as many families of meta-mappings over tasks (for example, multiplying polynomials by a constant, squaring them, or permuting their input variables). This allows us to not only examine the ability of the system to learn to learn polynomials from data, but also to adapt its learned representations in accordance with these meta-tasks. Details of the architecture and training can be found in appendix \ref{app_detailed_methods}.\par
%\vspace{-0.7em}
\paragraph{Basic meta-learning:} First, we show that the system is able to achieve the basic goal of learning a held-out polynomial from a few data points in fig. \ref{poly_basic_results} (with good sample-efficiency, see supp. fig. \ref{supp_poly_sweep_results}). \par 
%\vspace{-0.7em}
\paragraph{Meta-mapping (task $\rightarrow$ task):} Furthermore, the system is able to perform meta-mappings over polynomials in order to flexibly reconfigure its behavior (fig. \ref{poly_meta_map_results_examples}). We train the system to perform a variety of mappings, for example switch the first two inputs of the polynomial, add 3 to the polynomial, or square the polynomial. We then test its ability to generalize to held-out mappings, for example a held-out input permutation, or an unseen additive shift. The system is both able to apply learned meta-mappings to held-out basic tasks, and to apply held-out meta-mappings it has not been trained on, simply by seeing examples of the mapping. \par

@@

In order to adapt the HoMM architecture, we made a very simple change. Instead of providing the system with (input, target) tuples to embed, we provided it with (state, action, reward) tuples, and trained it via a DQN-like approach \citep{Mnih2015} to predict rewards for each bet in each state. (A full RL framework is not strictly necessary here because there is no temporal aspect to the tasks; however, because the outcome is only observed for the action you take, it is not a standard supervised task.) The hand is explicitly provided to the network for each example, but which game is being played is implicitly captured in the training examples, without any explicit cues. That is, the system must learn to play directly from seeing a set of (state, action, reward) tuples which implicitly capture the structure and stochasticity of the game. We also trained the system on meta-tasks, such as to identifying the various types of games and options, and to make meta-mappings, for example switching from trying to win a game to trying to lose. Details of the architecture and training can be found in appendix \ref{app_detailed_methods}. \par

@@

\paragraph{Future Directions:} We think that the general perspective of embedding tasks and considering meta-mappings will yield many fruitful future directions. We hope that our work will inspire more exploration of behavioral adaptation, in areas beyond the simple domains we considered here. To this end, we suggest the creation of meta-learning datasets which include information not only about tasks, but about the relationships between them. For example,  as can reinforcement learning tasks which involve executing instructions \citep[e.g.][]{Hermann2017, Co-Reyes2019} can be usefully interpreted from this perspective. Furthermore, we think our work provides a novel perspective on the types of flexibility that human intelligence exhibits, and thus hope that it may have implications for cognitive science. \par 

>>>>

%\vspace{-0.25em}
\section{Homoiconic meta-mapping (HoMM) architecture}
%\vspace{-0.5em} % section
To address these challenges, we propose HoMM architectures, composed of two components: 
\vspace{-0.5em}
\begin{enumerate} \setlength \itemsep{0em}
\item Input/output systems: domain specific encoders and decoders (vision, language, etc.) that map into a shared embedding space $Z$.
\item A meta-learning system that a) learns to embed tasks into the shared embedding space $Z$, b) learns to use these task embeddings to perform task-appropriate behavior, c) learns to embed meta-mappings into the same space, and d) learns to use these meta-mapping embeddings to transform basic task embeddings in a meta-mapping appropriate way.\end{enumerate}
\vspace{-0.5em}
\looseness=-1
These architectures are homoiconic because they have a completely shared $Z$ for individual data points, tasks, and meta-mappings. Why is this useful? The primary advantage is that it parsimoniously allows for arbitrary mappings between these entities. In addition to basic tasks, the system can learn to perform meta-mappings to follow instructions or change behavior. That is, it can transform task representations using the same components it uses to transform basic data points. (See also appendix \ref{app_lesion_results_shared_z}.) \par
Without training on meta-mappings, of course, the system will not be able to execute them well. However, as we will show, if it is trained on a broad enough set of such mappings, it will be able to generalize to new instances drawn from the same meta-mapping distribution. For instances that fall outside its data distribution, or for optimal performance, it may require some retraining, however. This reflects the structure of human behavior -- we are able to adapt rapidly when new knowledge is relatively consistent with our prior knowledge, but learning an entirely new paradigm (such as calculus for a new student) can be quite slow \citep[cf.][]{Kumaran2016, Botvinick2019}. \par 
\begin{figure}[t]
\centering
\begin{tikzpicture}[scale=0.8, every node/.style={scale=0.8}]
%% basic meta learning
\begin{scope}[shift={(0.4, 0.4)}]
\draw[boundingbox, fill=white] (-3, -4.3) rectangle (2.5, 3.2);
\end{scope}
\begin{scope}[shift={(0.2, 0.2)}]
\draw[boundingbox, fill=white] (-3, -4.3) rectangle (2.5, 3.2);
\end{scope}

@@

\caption{The HoMM architecture allows for transformations at different levels of abstraction. (a) For basic meta-learning a dataset consisting of (input embedding, output embedding) tuples is processed by the meta-network \(\mathcal{M}\) to produce a function embedding \(z^{func}\), which is processed by the hyper network \(\mathcal{H}\) to parameterize a function \(F_{z^{func}}\), which attempts to compute the transformation on held-out probe inputs. However, our approach goes beyond basic meta-learning. The function embedding \(z^{func}\) can then be seen as a single input or output at the next level of abstraction, when the same networks \(\mathcal{M}\) and \(\mathcal{H}\) are used to transform function embeddings based on examples of a meta-mapping (b). To evaluate meta-mapping performance, a probe embedding of a held-out function is transformed by the architecture to yield a predicted embedding for the transformed task. The performance of this predicted embedding is evaluated by moving back down a level of abstraction and evaluating on the actual target task (c). Because the function embedding is predicted by a transformation rather than from examples, new tasks can be performed zero-shot. (\(M\) and \(H\) are learnable deep networks, and $F_{z}$ is a deep network parameterized by $\mathcal{H}$ conditioned on function embedding \(z\). Input and output encoders/decoders are omitted for simplicity. See the text and appendix \ref{app_model_details} for details.)} \label{architecture_inference_fig}

@@

\vspace{-0.5em}
\paragraph{Basic meta-learning:} To perform a basic task, input and target encoders ($\mathcal{I}$ and $\mathcal{T}$) are used to embed individual pairs from an example dataset \(D_1\), to form a dataset of example (input, output) tuples (fig. \ref{architecture_inference_fig}a). These examples are fed to $\mathcal{M}$, which produces a function embedding (via a deep neural network, with several layers of parallel processing across examples, followed by an element-wise max across examples, and several more layers). This function embedding is mapped through the hyper network $\mathcal{H}$ to parameterize $F$, and then $F$ is used to process a dataset of embedded probe inputs, and $\mathcal{O}$ to map the resultant embeddings to outputs. This system can be trained end-to-end on target outputs for the probes. Having two distinct datasets forces generalization at the meta-learning level, see appendix \ref{app_clarifying_holdouts}. See appendix \ref{app_model_details} for detailed architecture, and hyper-parameters. \par 

@@

\vspace{-0.25em}

@@

\begin{figure}[t]
\centering
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/poly/meta_results.png}
\caption{The polynomials domain, section \ref{sec_poly}.}
\label{poly_meta_map_results_examples}
\end{subfigure}%
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{figures/meta_mapping.png}
\caption{The cards domain, section \ref{sec_cards}.}
\label{cards_meta_map_results_examples}
\end{subfigure}%
\caption{The HoMM architecture performs well at meta-mappings. (\subref{poly_meta_map_results_examples}) The system generalizes to apply learned meta-mappings to new polynomials, and even to apply unseen meta-mappings. The plots show the loss produced when evaluating the mapped embedding on the target task. For example, if the initial polynomial is $p(x) = x + 1$, and the meta-task is ``square,'' the loss would be evaluated by transforming the embedding of $p(x)$ and evaluating how well the mapped embedding regresses on to $p(x)^2 = x^2 + 2x + 1$. The results show that the system succeeds at applying meta-mappings it is trained on to held-out polynomials, as well as applying held-out meta-mappings to either trained or held-out polynomials. The solid line indicates optimal performance; the dashed line is untrained model performance.
(\subref{cards_meta_map_results_examples}) The system generalizes to meta-mapping new tasks in the cards domain. The system is trained to do the meta-mappings shown here on a subset of its basic tasks, and is able to generalize these mappings to perform novel tasks zero-shot. For example, for the ``losers'' mapping, the sytem is trained to map games to their losers variants. When given a held-out game, it is able to apply the mapping to guess how to play the losing variation. This plot shows the reward produced by taking the mapped embedding and playing the targeted game. The gray dashed line indicates random performance, while the colored dashed lines indicate performance if the system did not alter its behavior in response to the meta-mapping. The system generally exceeds these baselines, although the switch-suits baseline is more difficult with the targeted holdout. 
Error-bars are bootstrap 95\%-CIs.} 
\label{meta_map_results}
\end{figure}

As a proof of concept, we first evaluated the system on the task of learning polynomials of degree $\leq 2$ in 4 variables (i.e. the task was to regress functions of the form $p: \mathbb{R}^4 \rightarrow \mathbb{R}$ where $p \in \mathcal{P}_2 \left(\mathbb{R}\right)$, though the model was given no prior inductive bias toward polynomial forms). For example, if $p(w,x,y,z) = x$, the model might see examples like $(-1,1,1,1; 1)$ and $(0.7, 2.1, 1.3, -4; 2.1)$, and be evaluated on its output for points like $(-1, -1.3, 0.5, 0.3)$. This yields an infinite family of base-level tasks (the vector space of all such polynomials), as well as many families of meta-mappings over tasks (for example, multiplying polynomials by a constant, squaring them, or permuting their input variables). This allows us to not only examine the ability of the system to learn to learn polynomials from data, but also to adapt its learned representations in accordance with these meta-tasks. Details of the architecture and training can be found in appendix \ref{app_detailed_methods}.\par
\vspace{-0.7em}
\paragraph{Basic meta-learning:} First, we show that the system is able to achieve the basic goal of learning a held-out polynomial from a few data points in fig. \ref{poly_basic_results} (with good sample-efficiency, see supp. fig. \ref{supp_poly_sweep_results}). \par 
\vspace{-0.7em}
\paragraph{Meta-mapping (task $\rightarrow$ task):} Furthermore, the system is able to perform meta-mappings over polynomials in order to flexibly reconfigure its behavior (fig. \ref{poly_meta_map_results_examples}). We train the system to perform a variety of mappings, for example switch the first two inputs of the polynomial, add 3 to the polynomial, or square the polynomial. We then test its ability to generalize to held-out mappings from examples, for example a held-out input permutation, or an unseen additive shift. The system is both able to apply learned meta-mappings to held-out polynomials, and to apply held-out meta-mappings it has not been trained on, simply by seeing examples of the mapping. \par

@@

In order to adapt the HoMM architecture, we made a very simple change. Instead of providing the system with (input, target) tuples to embed, we provided it with (state, action, reward) tuples, and trained it to predict rewards for each bet in each state. (A full RL framework is not strictly necessary here because there is no temporal aspect to the tasks; however, because the outcome is only observed for the action you take, it is not a standard supervised task.) The hand is explicitly provided to the network for each example, but which game is being played is implicitly captured in the training examples, without any explicit cues. That is, the system must learn to play directly from seeing a set of (state, action, reward) tuples which implicitly capture the structure and stochasticity of the game. We also trained the system to make meta-mappings, for example switching from trying to win a game to trying to lose. Details of the architecture and training can be found in appendix \ref{app_detailed_methods}. \par

@@

\paragraph{Future Directions:} We think that the general perspective of considering meta-mappings will yield many fruitful future directions. We hope that our work will inspire more exploration of behavioral adaptation, in areas beyond the simple domains we considered here. To this end, we suggest the creation of meta-learning datasets which include information not only about tasks, but about the relationships between them. For example, reinforcement learning tasks which involve executing instructions \citep[e.g.][]{Hermann2017, Co-Reyes2019} can be usefully interpreted from this perspective. Furthermore, we think our work provides a novel perspective on the types of flexibility that human intelligence exhibits, and thus hope that it may have implications for cognitive science. \par 
